Multiple Dice Working as One: CAD Flows and Routing Architectures for Silicon Interposer FPGAs

### 2014-12-24 doc/journal/fpgapaperfinal/FPGAPaper/partitioning.tex
More wordsmithing
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4558 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\begin{enumerate}
  \item Leaving both the packer and the placer unconstrained. This is the base flow that does not use partitioning, as described in Section \ref{cadSection}.
  \item Partitioning the primitive netlist to guide the packer, leaving the placer unconstrained.
  \item Leaving the packer unconstrained, and partitioning the function block (clustered) netlist to guide the placer.
  \item Partitioning the primitive netlist to guide the packer, and constraining the placer based on constraints derived from partitioning, as described above.
\end{enumerate}

\subsubsection{Partitioning Tool}
Two well-known freely available graph partitioning tools are Metis~\cite{karypis1998multilevelmetis} and hMetis~\cite{karypis1999multilevelhmetis}. Metis operates on graphs, while hMetis operates on hypergraphs -- a hypergraph is a generalization of a graph, where each edge can connect any number of vertices. It is most natural to represent a circuit netlist as a hypergraph, where the vertices are netlist pins (or blocks) and the hyperedges are the nets joining the pins. Based on that criterion alone, hMetis would be the most appropriate choice of partitioner. However, there are other important criteria to consider for our use case.

Graph partitioning algorithms generate partitions of a circuit while satisfying ``balance constraints''. The balance constraints exist to drive the algorithm away from the trivial partitioning solution where all vertices are in the same partition, leaving all other partitions empty. An example of a basic balance constraint is one which requires an equal number of vertices in each partition. Generally such a perfect balance overly restricts the solution, so some fractional ``unbalance'' is allowed; in this work we use $0.05$ ($5\%$). For two partitions, $P1$ and $P2$, this translates to~\cite{karypismanual}

@@

To derive an appropriate balance constraint for our use case, consider that FPGA circuit netlists may contain different block types such as LUTs, registers, RAM slices, and multipliers, among others. The basic balance constraint of (\ref{eq:balance_constraint}), applied to the two partition case with only LUTs and multipliers would ensure that the total block count (sum of the LUT count + multiplier count) in each partition is roughly balanced.

The problem with (\ref{eq:balance_constraint}) in the case of multiple block types is that the capacity of each partition is not taken into account on a per-block-type basis. It is therefore possible to generate partitions that cannot be realized because the number of resources of each type is not balanced across partitions, even though the total number of resources is balanced. For example, there is nothing preventing the partitioner from moving all LUTs to partition 1 and all multipliers to partition 2, which would not be physically realizable because partition 1 may not have enough LUTs to implement the desired partitioning (and no LUTs in partition 2 would be used.) One solution is to enforce a balance constraint on each block type, $i$:

\begin{equation}
\forall i, \abs*{ \frac{| P1_i |}{|P1_i \cup P2_i |} - 0.5} < unbalance
\end{equation}
where $P1_i$ is the set of all blocks of type $i$ in partition 1.

We refer to this type of per-block set of constraints as heterogeneous balance constraints. Note that there are complex legality constraints governing which LUTs and FFs can be packed into legal logic blocks, as well as which multipliers can be packed into legal DSP blocks~\cite{luu2014vtr}. As the partitioner is not aware of these constraints, flow 3 above (which runs the partitioner after function block packing) will be able to more precisely balance resource use across partitions.

The latest public version of hMetis does not support heterogeneous balance constraints, though Metis does. Metis was used in our CAD flow. However, as discussed earlier, the circuit netlist is most naturally represented as a hypergraph, and so it needs to be transformed to a graph before Metis can process it.

@@

We explored several ways to generate the per-net subgraphs. The two parameters in this operation are the graph topology and the edge weight scheme. We considered two subgraph topologies, clique and star, as illustrated in Figure \ref{fig:star_clique}. We assigned the same edge weight to every edge of the generated subgraph. Edge weights were computed based on the number of vertices in the hyperedge, $n$, and we considered edge weights equal to $1$, $1/n$, and $1/n^2$.

@@

The parameters of graph topology and edge weight were evaluated by generating the total primitive netlist graph and running it through the Metis partitioner to generate two partitions. The results of the different schemes were ranked based on the hyperedge cutsize, which is the number of hyperedges crossing the cutline, as illustrated in Figure \ref{fig:hyperedge_cutline}. The hyperedge cutsize is a proxy for the routability of the circuit, as the cutsize is the number of nets that cross the interposer boundary. To compute the hyperedge cutsize, we applied the partitioning result (which assigns each vertex to a partition number) to the original untransformed hypergraph, and counted the number of hyperedges whose vertices were spread across different partitions. We computed the hyperedge cutsize for every combination of graph topology and graph edge weight scheme, across several circuits. The results are shown in Figure \ref{fig:graph_topology_cutsize}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{hyperedge_cutline.pdf}
\caption{A graph with six vertices and three hyperedges partitioned into two halves. As there are two hyperedges crossing the cutline, the hyperedge cutsize is 2.}
\label{fig:hyperedge_cutline}
\end{figure}

\begin{figure}[!htbp]
\centering
%\includegraphics[width=\linewidth]{graph_topology_cutsize.eps}
%\input{graph_topology_cutsize.tex}
\input{graph_topology_cutsize}
\caption{Hypergraph cut size achieved by Metis with different hyperedge models and weightings. n is the number of vertices on a hyperedge.}
\label{fig:graph_topology_cutsize}
\end{figure}

@@

For both topologies, the $1/n$ edge weight scheme gives the smallest cut size on average. This can be explained intuitively in terms of the total edge weight over all edges of the graph. The star $1$ (constant) edge weight scheme assigns a total weight of $n$ to each net, which heavily weights high-fanout nets. In contrast, the star $1/n$ scheme assigns the same total weight of $1$ to each net. The star $1/n^2$ scheme penalizes high-fanout nets relative to lower-fanout nets. Since we seek to minimize the hyperedge cut, it makes intuitive sense to weight all hyperedges equally, and the star $1/n$ scheme achieves this.

To validate the choice of hyperedge cutsize as a proxy for circuit routability, we ran the partitioning CAD flow for each graph topology and edge weight scheme, across several circuits. We use an unbalance of $5\%$, split the clustered netlist into two partitions (constraining only the placer), cut $80\%$ of the wires crossing the interposer, and impose a $1ns$ delay penalty for wires crossing the interposer. To accommodate unbalance in the placement engine, we increase the size of the grid of complex blocks that make up the FPGA device. Relative to the minimum device size required for placement (without partitioning constraints), we add $10\%$ to the complex block grid width and $10\%$ to the grid height. Figure \ref{fig:graph_topology_mcw} shows the geometric mean of the minimum channel width required for a successful route, for each of the graph topology and edge weight combinations.

@@

The star topology with 1/n edge weights achieves the best minimum channel width, confirming that hyperedge cutsize is a good proxy for routability. Consequently, we use this scheme in all future results in this paper. Unlike routability, we found that the post-routing critical path delay was not strongly impacted by the hypergraph conversion scheme used.

@@

\begin{figure}[!htbp]
\centering
%\includegraphics[width=\linewidth]{vpr_flows_mcw.eps}
\input{vpr_flows_mcw}
\caption{Minimum channel width achieved by VPR with the different partitioning flows.}
\label{fig:flows_mcw}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\input{vpr_flows_crit_path}
\caption{Critical path delay achieved by VPR with the different partitioning flows.}
\label{fig:flows_crit_path}
\end{figure}

@@

Considering both routability and delay, the best CAD flow constrained both packing and placement with the partitioning result and this is the flow we use in the remainder of this paper.

\subsubsection{\% wires cut}
To see how well the CAD flow works for a range of interposer architectures, we varied the fraction of wires cut and measured the average minimum channel width over the 8 largest VTR circuits. As shown in Figure \ref{fig:wires_cut}, this new flow is a significant improvement over that of [] for FPGAs where more than half the wires are cut at interposer boundaries. It is slightly worse than the traditionally FPGA CAD flow even for an architecture with 0\% of the wires cut, which is a monolithic FPGA, presumably because we have restricted the placer's freedom.

>>>>

\begin{enumerate}
  \item Leaving both the packer and the placer unconstrained. This is the base flow that does not use partitioning, as described in Section \ref{cadSection}.
  \item Partitioning the primitive netlist to guide the packer, leaving the placer unconstrained.
  \item Leaving the packer unconstrained, and partitioning the function block (clustered) netlist to guide the placer.
  \item Partitioning the primitive netlist to guide the packer, and, as described above, passing these constraints to the placer.
\end{enumerate}

\subsubsection{Partitioning Tool}
Two well-known freely available graph partitioning tools are Metis~\cite{karypis1998multilevelmetis} and hMetis~\cite{karypis1999multilevelhmetis}. Metis operates on graphs, while hMetis operates on hypergraphs -- a hypergraph is a generalization of a graph, where each edge can connect any number of vertices. It is most natural to represent a circuit netlist as a hypergraph, where the vertices are netlist pins (or blocks) and the hyperedges are the nets joining the pins. Owing to the natural hypergraph representation, hMetis would be the most appropriate choice of partitioner. However, there are other issues to consider for our use case.

Graph partitioning algorithms assign vertices to partitions while satisfying ``balance constraints''. The balance constraints exist to drive the algorithm away from the trivial partitioning solution where all vertices are in the same partition, leaving all other partitions empty. An example of a basic balance constraint is one which requires an equal number of vertices in each partition. Generally such a perfect balance overly restricts the solution, so some fractional ``unbalance'' is allowed; in this work we use $0.05$ ($5\%$). For two partitions, $P1$ and $P2$, this translates to~\cite{karypismanual}

@@

To derive an appropriate balance constraint for our use case, consider that FPGA circuit netlists may contain different block types such as LUTs, registers, RAM slices, and multipliers, among others. The basic balance constraint of (\ref{eq:balance_constraint}), applied to a netlist with only LUTs and multipliers, would ensure that the total block count (sum of the LUT count + multiplier count) in each partition is roughly balanced. The problem with (\ref{eq:balance_constraint}) in the case of multiple block types is that the capacity, per block type, of each partition is not taken into account. It is therefore possible to generate partitions that cannot be realized because the number of resources of each type is not balanced across partitions, even though the total number of resources is balanced. For example, there is nothing preventing the partitioner from moving all LUTs to partition 1 and all multipliers to partition 2, which may not be realizable given the per-block-type capacity of each partition.

One solution is to enforce a balance constraint on each block type, $i$:

\begin{equation}
\forall i, \abs*{ \frac{| P1_i |}{|P1_i \cup P2_i |} - 0.5} < unbalance
\end{equation}
where $P1_i$ is the set of all blocks of type $i$ in partition 1. We refer to this type of per-block set of constraints as heterogeneous balance constraints. Note that there are complex legality constraints governing which LUTs and FFs can be packed into legal logic blocks, as well as which multipliers can be packed into legal DSP blocks~\cite{luu2014vtr}. As the partitioner is not aware of these constraints, flow 3 above (which runs the partitioner after function block packing) will be able to more precisely balance resource use across partitions.

The current version of hMetis does not support heterogeneous balance constraints, though Metis does. Metis was used in our CAD flow. However, as the circuit netlist is naturally represented as a hypergraph, it needs to be transformed to a graph before Metis can process it.

@@

We explored several ways to generate the per-net subgraphs. The two parameters in this operation are the graph topology and the edge weight scheme. We considered two graph topologies, clique and star, as illustrated in Figure \ref{fig:star_clique}. We assigned the same edge weight to every edge of the generated subgraph. Edge weights were computed based on the number of vertices in the originating hyperedge, $n$, and we considered edge weights equal to $1$, $1/n$, and $1/n^2$. We use \emph{hyperedge model} to refer to the combination of graph topology and edge weight scheme.

@@

The number of hyperedges crossing the cutline is referred to as the \emph{hyperedge cutsize}, as illustrated in Figure \ref{fig:hyperedge_cutline}. The hyperedge cutsize is a measure of the routability of the circuit, as it equals the number of nets that cross the interposer boundary. Using the Metis partitioner to generate graph partitions (which assigns each vertex to a partition number) and applying the partitioning result to the original untransformed hypergraph, we computed the hyperedge cutsize for every hyperedge model, across several circuits. We used the results, shown in Figure \ref{fig:graph_topology_cutsize}, to select the ``best'' hyperedge model.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{hyperedge_cutline.pdf}
\caption{A graph with six vertices, three hyperedges, and two partitions. As there are two hyperedges crossing the cutline, the hyperedge cutsize is 2.}
\label{fig:hyperedge_cutline}
\end{figure}

\begin{figure}[!htbp]
\centering
%\includegraphics[width=\linewidth]{graph_topology_cutsize.eps}
%\input{graph_topology_cutsize.tex}
\input{graph_topology_cutsize}
\caption{Hypergraph cut size achieved by Metis with different hyperedge models and weightings. $n$ is the number of vertices on a hyperedge.}
\label{fig:graph_topology_cutsize}
\end{figure}

@@

For both topologies, the $1/n$ edge weight scheme gives the smallest cut size on average. This can be explained intuitively in terms of the total edge weight over all edges of the graph. The star $1$ (constant) edge weight scheme assigns a total weight of $n$ to each net, which heavily weights high-fanout nets. In contrast, the star $1/n$ model assigns the same total weight of $1$ to each net. The star $1/n^2$ model penalizes high-fanout nets relative to lower-fanout nets. Since we seek to minimize the hyperedge cut, it makes intuitive sense to weight all hyperedges equally, and the star $1/n$ model achieves this.

To validate the choice of hyperedge cutsize as a proxy for circuit routability, we ran the partitioning CAD flow for each hyperedge model, across several circuits. We use an unbalance of $5\%$, split the clustered netlist into two partitions (constraining only the placer), cut $80\%$ of the wires crossing the interposer, and impose a $1ns$ delay penalty for wires crossing the interposer. To accommodate unbalance in the placement engine, we increase the size of the grid of complex blocks that make up the FPGA device. Relative to the minimum device size required for placement (without partitioning constraints), we add $10\%$ to the complex block grid width and $10\%$ to the grid height. Figure \ref{fig:graph_topology_mcw} shows the geometric mean of the minimum channel width required for a successful route, for each hyperedge model.

@@

The star topology with 1/n edge weights achieves the best minimum channel width, confirming that hyperedge cutsize is a good proxy for routability. Consequently, we use this model in all future results in this paper. Unlike routability, we found that the post-routing critical path delay was not strongly impacted by the hypergraph model used.

@@

\begin{figure}[!htbp]
\centering
%\includegraphics[width=\linewidth]{vpr_flows_mcw.eps}
\input{vpr_flows_mcw}
\caption{Minimum channel width achieved by VPR for each partitioning flow.}
\label{fig:flows_mcw}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\input{vpr_flows_crit_path}
\caption{Critical path delay achieved by VPR for each partitioning flow.}
\label{fig:flows_crit_path}
\end{figure}

@@

Considering both routability and delay, the best CAD flow used the partitioning result to constrain both packing and placement, and we use that flow in the remainder of this paper.

\subsubsection{\% wires cut}
To see how well the CAD flow works for a range of interposer architectures, we varied the fraction of wires cut and measured the average minimum channel width over the 8 largest VTR circuits. As shown in Figure \ref{fig:wires_cut}, this new flow is a significant improvement over that of~\cite{interposer2014} for FPGAs where more than half the wires are cut at interposer boundaries. It is slightly worse than the traditional FPGA CAD flow even for an architecture with 0\% of the wires cut, which is a monolithic FPGA, presumably because we have restricted the placer's freedom.
